Synthesis of Linear Functional Programs

### 2021-07-22 article/article.tex
explain -oL, do todos, tweaks

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a system of
resource management~\cite{resman} and focusing.
%
% TODO: Onde falamos de resource management? ( 
%
In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Asynchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Synchronous connectives are dual. \todo{Say which
  are which}

@@

The core language is a simply-typed linear lambda calculus with linear
functions ($\lolli$), additive ($\with$) and multiplicative ($\tensor$) pairs,
multiplicative unit (\textbf{1}), additive sums ($\oplus$)
and the exponential modality ($\bang$) (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais? NAO :)
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (%heavily
inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
rule ($\lolli R$) is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

When we reach a non-invertible proposition on the right, we start inverting the
$\Omega$ context. The rule to transition to inversion on the left is:
\begin{mathpar}
    \infer*[right=($\Uparrow R$)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

@@

When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
we simply move it to the linear context $\Delta$, and keep inverting on the left:
\begin{mathpar}
    \infer*[right=($\Uparrow L$)]
    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
    \textrm{not left asynchronous}}
    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
\end{mathpar}

@@

\todo[inline]{Aqui tens de explicar uma regra pelo menos}
The decision rules are followed by either left or right focus rules:
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

@@

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}

where the ADT $T$ and its constructors stand for any ADT with form
\begin{minted}{haskell}
    data T = C1 X1 | C2 X2 | ... | Cn Xn
\end{minted}

@@

\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        T \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
    T \and T \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra
%
The rules above ensure that a recursive ADT will be deconstructed
once, and that subsequent equal ADTs will only be useable from the linear
context -- essentially forcing it to be used to instantiate another proposition,
which will typically be an argument for the recursive call.

@@

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
variables that stand for any (non-polymorphic) type in $\tau$. Such a type is
also called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into
a non-quantified type, and then treating its type variables uniformly.  First,
type variables are considered \emph{atomic types}, then, we instantiate the
bound variables of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
this, where $\forall \overline{\alpha}.\ \tau \sqsubseteq \tau'$ indicates type
$\tau'$ is an \emph{instantiation} of type scheme $\forall \overline{\alpha}.\
\tau$:
%
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
%
As such, the construction of a derivation in which the only rule that
can derive an atom is the \textsc{init} rule corresponds to the
synthesis of a program where some expressions are treated agnostically
-- nothing constrains their type -- i.e.~the program is
polymorphic. The simplest example is the polymorphic function
\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The
program synthesized from that specification is $\lambda x . x$, a
lambda abstraction that does not constrain the type of its parameter
$x$ in any way.
% TODO: \todo{R : rewrite this? não estranho?}

@@

The \synname\ synthetiser operates as part of the pipeline that processes a
full program in the \synname\ language, and is called to generate terms marked
(by the syntax \emph{\{\{ ... \}\}}) for
synthesis. The main pipeline consists of:
\[
    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow \textrm{Inference} \rightarrow \textrm{Synthesis}
\]

@@

The synthesis module iterates over the synthesis marks and runs the
synthetiser with its context and goal for each mark, minimizing the
resulting expression (i.e.~$\eta$-reducing it), and replacing the mark
in the frontend AST with the synthed program.

@@

\mypara{Memoization} Some proofs attempt to synthetise the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthetise result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values and
uses the result as the key\todo{via a hashconsing library?}.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We capture failure of a LogicT computation using the $\mathit{ifte}$
function and represent failure or success information with the \texttt{Maybe}
type. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take
around $5s$, while synthesis without memoization doesn't terminate
even after $15m$ running.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some
examples. We leave such optimizations to future work.

>>>>

\mypara{Core Rules} The core system comprises of proof search for
(intuitionistic) linear logic sequent calculus proofs, based on a system of
resource management~\cite{resman} and focusing.
%
% TODO: Onde falamos de resource management? ( 
%
In intuitionistic sequent calculi, each connective has a so-called \emph{left}
and a \emph{right} rule, which effectively define how to decompose an ambient
assumption of a given proposition and how to prove a certain proposition is
true, respectively.  Andreoli's \emph{focusing} for linear logic~\cite{Andreolli
focusing} is a technique to remove non-essential nondeterminism from
proof-search by structuring the application of so-called invertible and
non-invertible inference rules. Andreoli observed that the
connectives of linear logic can be divided into two categories, dubbed
synchronous and asynchronous. Asynchronous connectives are those whose
right rules are \emph{invertible}, i.e. they can be applied eagerly
during proof search without losing provability and so the order in
which these rules are applied is irrelevant, and whose left rules
are not invertible. Synchronous connectives are dual. The asynchronous \todo{Escrever isto depois
de apresentar o linear $\lambda$-calculus?} 
connectives are $\lolli$ and $\with$ and the synchronous are
$\tensor,\textbf{1},\oplus,\bang$.

@@

The core language is a simply-typed linear $\lambda$-calculus with linear
functions ($\lolli$), additive ($\with$) and multiplicative ($\tensor$) pairs,
multiplicative unit (\textbf{1}), additive sums ($\oplus$)
and the exponential modality ($\bang$) (to internalize unrestricted use of variables), for
which the terms ($M,N$) and types' ($\tau$) syntax is constructed with the
following grammar:
%
% TODO: Nao é um problema eu dizer que o A e B são átomos ali em cima e aqui
% usá-los como tipos gerais? NAO :)
\[
\begin{tabular}{lclc}
$M,N$ & $\ ::=\ $ & $u, v$ & \emph{variables} \\
    & & $\vert\ \lambda x. M\ \vert\ M\ N$ & $\lolli$ \\
    & & $\vert\ M \with N\ \vert\ \mathsf{fst}\ M\ \vert\ \mathsf{snd}\ M$ & $\with$ \\
    & & $\vert\ M \tensor N\ \vert\ \llet{u\tensor v = M}{N}$ & $\tensor$ \\
    & & $\vert\ \star\ \vert\ \llet{\star = M}{N}$ & $\textbf{1}$ \\
    & & $\vert\ \mathsf{inl}\ M\ \vert\ \mathsf{inr}\ M\ \vert\ (\mathsf{case}\
    M\ \mathsf{of\ inl}\ u \Rightarrow N_1\ \vert\ \mathsf{inr}\ v \Rightarrow
    N_2)$ & $\oplus$ \\
    & & $\vert\ \bang M\ \vert\ \llet{\bang u = M}{N}$ & $\bang$ \\
    \\
$\tau$ & $\ ::=\ $ & $A, B, C\ \vert\ \tau \lolli \tau\ \vert\ \tau \with \tau\
    \vert\ \tau \tensor \tau\ \vert\ \textbf{1}\ \vert\ \tau \oplus \tau\ \vert\ \bang \tau$
\end{tabular}
\]
%
Putting together linear logic and linear lambda calculus through the Curry-Howard correspondence, resource
management, and focusing, we get the following core formal system (%heavily
inspired
by Frank Pfenning's notes on focusing for linear logic~\cite{FPnotes...}) -- in which the
rule $\lolli$R is read: to synthesize a program of type $A \lolli B$ while inverting
right (the $\Uparrow$ on the goal), with unrestricted context $\Gamma$, linear
context $\Delta$, and inversion context $\Omega$, synthetise a program of type $B$ with
an additional hypothesis of type $A$ named $x$ in the $\Omega$ context,
resulting in the program $M$ and output linear
context $\Delta'$ that cannot contain the added hypothesis $x{:}A$. Finally, the
resulting program is $\lambda x . M$ and the output linear context is
$\Delta'$.

@@

When we reach a non-invertible proposition on the right, we start inverting the
$\Omega$ context. The rule to transition to inversion on the left is:
\begin{mathpar}
    \infer*[right=($\Uparrow$R)]
    {\Gamma ; \Delta/ \Delta' ; \Omega \Uparrow\ \vdash C \and C\ \textrm{not
    right asynchronous}}
    {\Gamma ; \Delta/\Delta' ; \Omega \vdash C \Uparrow}
\end{mathpar}

@@

When we find a synchronous (i.e. non-invertible) proposition in $\Omega$,
we simply move it to the linear context $\Delta$, and keep inverting on the left:
\begin{mathpar}
    \infer*[right=($\Uparrow$L)]
    {\Gamma; \Delta, A/\Delta'; \Omega \Uparrow\ \vdash C \and A\ 
    \textrm{not left asynchronous}}
    {\Gamma; \Delta/\Delta'; \Omega, A \Uparrow\ \vdash C}
\end{mathpar}

@@

The decision rules are followed by either left or right focus rules. The left
focus rule $\lolli$L is read ``to synthesize a program of type $C$ with input
linear context $\Delta$ and while left focused on a proposition $x$ of type $A
\lolli B$, synthesize a program $M$ of type $C$ with input $\Delta$ and 
left focused ($\Downarrow$) on $y$ of type
$B$, and output the context $\Delta'$ (with the unused
linear propositions); then, switching to the right inversion judgment ($\Uparrow$), synthesize a program $N$ of type $A$ with input
$\Delta'$ and output context $\Delta''$; finally, replace in $M$ all occurrences
of variable $y$ with program $x\ N$ (apply $x$ to $N$)''. It can be understood
as the synthesis of a program of type $C$ using the return value (of type $B$) from the function $A \lolli B$, and then the
synthesis of an argument of type $A$ to apply the function to.
\begin{mathpar}
    \infer*[right=($\tensor R$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow \and \Gamma ; \Delta'/\Delta'' \vdash N
    : B \Downarrow}
    {\Gamma; \Delta/\Delta'' \vdash (M \tensor N) : A \tensor B \Downarrow}
\and
    \infer*[right=($1 R$)]
    { }
    {\Gamma; \Delta/\Delta \vdash \star : \textbf{1} \Downarrow}
\and
    \infer*[right=($\oplus R_1$)]
    {\Gamma; \Delta/\Delta' \vdash M : A \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inl}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\oplus R_2$)]
    {\Gamma; \Delta/\Delta' \vdash M : B \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ \textrm{inr}\ M : A \oplus B \Downarrow}
\and
    \infer*[right=($\bang R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash M : A \Uparrow \and \Delta = \Delta'}
    {\Gamma; \Delta/\Delta \vdash \bang M : \bang A \Downarrow}
\and
    \infer*[right=($\lolli L$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C \and \Gamma;
    \Delta'/\Delta''; \cdot \vdash N : A \Uparrow}
    {\Gamma; \Delta/\Delta''; x{:}A \lolli B \Downarrow\ \vdash M\{(x\,N)/y\} : C}
\and
    \infer*[right=($\with L_1$)]
    {\Gamma; \Delta/\Delta'; y{:}A \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{fst}\ x)/y\} : C}
\and
    \infer*[right=($\with L_2$)]
    {\Gamma; \Delta/\Delta'; y{:}B \Downarrow\ \vdash M : C}
    {\Gamma; \Delta/\Delta'; x{:}A \with B \Downarrow\ \vdash M\{(\textrm{snd}\ x)/y\} : C}
\end{mathpar}
%
Eventually, the focus proposition will no longer be synchronous, i.e. it's
atomic or asynchronous. If we're left focused on an atomic proposition we either
instantiate the goal or fail. Otherwise the left focus is asynchronous and we
can start inverting it. If we're right focused on a proposition that isn't right
synchronous, we switch to inversion as well. Three rules model these conditions:
%
\begin{mathpar}
    \infer*[right=(init)]
    {  }
    {\Gamma; \Delta/\Delta'; x{:}A \Downarrow\ \vdash x : A}
\and
    \infer*[right=($\Downarrow L$)]
    {\Gamma; \Delta/\Delta'; A \Uparrow\ \vdash C \and A\ \textrm{not atomic and not left synchronous}}
    {\Gamma; \Delta/\Delta'; A \Downarrow\ \vdash C}
\and
    \infer*[right=($\Downarrow R$)]
    {\Gamma; \Delta/\Delta'; \cdot \vdash A \Uparrow}
    {\Gamma; \Delta/\Delta' \vdash A \Downarrow}
\end{mathpar}
The rules written above together make the core of our synthetiser. Next, we'll
present new rules that align and build on top of these to synthetise recursive
programs from more expressive (richer) types.

@@

\begin{mathpar}
    \infer*[right=(adtR)]
    {\Gamma; \Delta/\Delta' \vdash M : X_n \Downarrow}
    {\Gamma; \Delta/\Delta' \vdash\ C_n \ M : T \Downarrow}
\and
    \mprset{flushleft}
    \infer*[right=(adtL)]
    {
        \Gamma ; \Delta/ \Delta'_1 ; \Omega, y_1{:}X_1 \Uparrow\ \vdash M_1 : C
        \and
        y_1 \notin \Delta'_1
        \\
        \Gamma ; \Delta/ \Delta'_2 ; \Omega, y_2{:}X_2 \Uparrow\ \vdash M_2 : C
        \\
        y_2 \notin \Delta'_2
        \\\\
        \\ \dots
        \\
        \Gamma ; \Delta/ \Delta'_n ; \Omega, y_n{:}X_n \Uparrow\ \vdash M_n : C
        \\
        y_n \notin \Delta'_n
        \\
        \Delta'_1 = \Delta'_2 = \dots = \Delta'_n
    }
    {\Gamma ; \Delta/\Delta'_1 ; \Omega, x{:}T \Uparrow\
    \vdash\ \textrm{case}\ x\ \textrm{of}\ \dots\ |\ C_n\ y_n
    \rightarrow M_n : C}
\end{mathpar}
where the ADT $T$ and its constructors stand for any ADT with form \begin{minted}{haskell}
    data T = C1 X1 | C2 X2 | ... | Cn Xn
\end{minted}

@@

\begin{mathpar}
    \infer*[right=(adt$\Uparrow$L)]
    {
        (\Rho_C; \Rho_D);\Gamma; \Delta, x{:}T/\Delta'; \Omega \Uparrow\ \vdash M : C
        \and
        T \in \Rho_D
    }
    {(\Rho_C; \Rho_D);\Gamma; \Delta/\Delta'; \Omega, x{:}T \Uparrow\ \vdash M : C}
    \and
    \infer*[right=(adt-init)]
    {  }
    {\Rho; \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash x : T}
    \and
    \infer*[right=(adt$\Downarrow$L)]
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Uparrow\ \vdash M :
    T \and T \notin \Rho_D}
    {(\Rho_C; \Rho_D); \Gamma; \Delta/\Delta'; x{:}T \Downarrow\ \vdash M : T}
\end{mathpar}
% TODO: Adicionar regra que diz que se for ADT Rec -o A então a construção de
% ADT Rec é logo restrita?? parece que acelera minimamente mas não arranjo um
% exemplo em que seja crucial. pode ser uma coisa que tenha ficado para resolver
% uma coisa antiga que ficou resolvida mais tarde com outra modificação e então
% é agora inutil. Talvez seja demais adicionar à regra
%
Altogether, the rules above ensure that a recursive ADT will be deconstructed
once, and that subsequent equal ADTs will only be useable from the linear
context -- essentially forcing them to be used to instantiate another proposition,
which will typically be an argument for the recursive call.

@@

\mypara{Polymorphic Types} A polymorphic specification is a type of form
$\forall \overline{\alpha}.\ \tau$ where $\overline{\alpha}$ is a set of
variables that stand for any (non-polymorphic) type in $\tau$. Such a type is
also called a \emph{scheme}.  Synthesis of a scheme comprises of turning it into
a non-quantified type, and then treating its type variables uniformly.  First,
type variables are considered \emph{atomic types}, then, we instantiate the
bound variables of the scheme as described by the Hindley-Milner inference
method's~\cite{HM-infer} instantiation rule (put simply, generate fresh names
for each bound type variable); e.g. the scheme $\forall \alpha.\ \alpha \lolli
\alpha$ could be instantiated to $\alpha0 \lolli \alpha0$. We add a rule for
this, where $\forall \overline{\alpha}.\ \tau \sqsubseteq \tau'$ indicates type
$\tau'$ is an \emph{instantiation} of type scheme $\forall \overline{\alpha}.\
\tau$:
%
\begin{mathpar}
    \infer*[right=($\forall R$)]
    { \Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \tau' \Uparrow \and \forall
    \overline{\alpha}.\ \tau
    \sqsubseteq \tau'}
    {\Rho; \Gamma; \Delta/\Delta'; \Omega \vdash \forall \overline{\alpha}.\
    \tau \Uparrow}
\end{mathpar}
%
As such, the construction of a derivation in which the only rule that can derive
an atom is the \textsc{init} rule corresponds to the synthesis of a program
where some expressions are treated agnostically (nothing constrains their type),
i.e.~a polymorphic program. The simplest example is the polymorphic function
\emph{id} of type $\forall \alpha .\ \alpha \lolli \alpha$. The program
synthesized from that specification is $\lambda x . x$, a lambda abstraction
that does not constrain the type of its parameter $x$ in any way.

@@

The \synname\ synthetiser operates as part of the pipeline that processes a
full program in the \synname\ language, and is called to generate terms marked
(by the syntax \emph{\{\{ ... \}\}}) for
synthesis. The main pipeline consists of:
\[
    \textrm{Parsing} \rightarrow \textrm{Desugaring} \rightarrow
    \textrm{Inference} \rightarrow \textrm{Synthesis} \rightarrow
    \textrm{Evaluation}
\]

@@

The synthesis module iterates over the synthesis marks and runs the synthetiser
with the respective context and goal for each mark. The resulting expression is checked
against the synthesis constraints defined through some keywords, continuing
synthesis if they are violated: ``using'' guarantees that certain functions are
used in the program body, and ``assert'' evaluates the given assertion within the
synthesized program to a validating boolean.  Additionally, the keyword
``depth'' controls the \emph{existential depth} (seen
in~\ref{sec:formal_system}), and ``choose'' selects a different result as long
as one is available. Finally, the expression is minimized (i.e.~$\eta$-reduced),
and the mark in the frontend AST is replaced with the synthesized program.

The evaluation module evaluates the function named ``main'' within the
complete synthesized program, returning a final value.

@@

\mypara{Memoization} Some proofs attempt to synthetise the same
sub-goal with the same premises more than once. Since these proof
attempts might be expensive performance-wise, and we can rely on the
program's determinism (equal calls to synthetise result in equal
outcomes), we added memoization to our synthesizer -- an optimization
technique that stores function calls and returns cached values for
equal inputs.
%
For our synthesizer, two calls are equivalent when
contexts $\Theta, \Rho, \Gamma, \Delta$, $\Omega$, (or alternatively
the left focus), and the goal, are equivalent. Because comparing all
of these is slow, our implementation hashes the relevant values via a hashing
library (\emph{Hashable}~\cite{https://hackage.haskell.org/package/hashable}) and
uses the result as the key.
%
Furthermore, a branch we want to memoize
might fail, not providing any result -- we also want to record
synthesis failure, as it happens more often than successful
results. We capture failure of a LogicT computation using the $\mathit{ifte}$
function and represent failure or success information with the \texttt{Maybe}
type. We ran benchmarks to measure the impact of memoization on 
performance. It's quite noticeable in synthesis with multiple
functions in the unrestricted context, or more complex
specifications. For example, if we add a new function \emph{read} to
the unrestricted context in the \emph{array} problem seen in
Section~\ref{sec:overview}, synthesis with memoization will take
around $5s$, while synthesis without memoization doesn't terminate
even after $15m$ running.
%
The implementation could be improved by better combining the
memoization with the LogicT monad transformer (remembering decisions across
backtracking branches) and by finetuning key-equality by ignoring context
components that aren't relevant to the synthesis call, to name some
examples. We leave such optimizations to future work.
